---
title: "Getting started with MetaHunt"
output:
  rmarkdown::html_vignette: default
  pdf_document:
    latex_engine: xelatex
    toc: true
    number_sections: false
vignette: >
  %\VignetteIndexEntry{Getting started with MetaHunt}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 6, fig.height = 4,
  fig.align = "center"
)
```

```{r setup}
library(MetaHunt)
set.seed(1)
```

## What MetaHunt does

You have several **studies** (trial centres, cohorts, sites). Each study
has produced a **fitted model** for some function of patient-level
covariates — a regression function, a CATE function, a dose–response
curve. You also have **study-level metadata** for each study (region,
sample composition, year, …).

You want to **predict that function for a new study** characterised only
by its metadata, with a 95% prediction band, *without ever pooling
patient-level data across studies*.

MetaHunt does this in three steps:

1. **Basis hunting** — recover a small set of latent basis functions
   such that every study's function lies (approximately) in the convex
   hull of those bases.
2. **Weight model** — model how the mixing weights over those bases
   depend on study-level metadata.
3. **Conformal prediction** — distribution-free prediction intervals
   for the new study's function.

This vignette walks through the full pipeline.

## What you need to provide

Two ingredients drive every function in the package:

| Object | Description | Example |
|---|---|---|
| `F_hat` | An `m × G` numeric matrix. Row `i` is the function of study `i` evaluated on a *shared* set of `G` patient-level grid points. | `F_hat[i, g]` is the predicted treatment effect for centre `i` at "patient profile" `g`. |
| `W`     | An `m × p` matrix or data frame of study-level covariates (one row per study). | Region, mean age, percent female, year. |

Many users start with **fitted model objects** (e.g. random forests),
not `F_hat`. The next section shows how to build `F_hat` from such a
list.

## Preparing your data: from fitted models to `F_hat`

Suppose each centre has shared its fitted model object with you, plus
some metadata. The two onramp helpers in MetaHunt let you turn this
into the `(F_hat, W)` pair that the rest of the package consumes.

### Step 1: choose a shared evaluation grid

The grid is a finite set of patient profiles at which every centre's
model will be evaluated. Use [`build_grid()`] to construct it from any
reference patient-level dataset:

```{r build-grid-demo}
# A reference patient-level dataset (e.g. a target-population sample,
# or a public dataset like NHANES, or a held-out validation cohort).
ref <- data.frame(age   = rnorm(500, 60, 10),
                  bp    = rnorm(500, 130, 15),
                  bmi   = rnorm(500, 28, 4))

grid <- build_grid(ref, n_grid = 50, seed = 1)
head(grid)
```

The grid does **not** have to come from any specific centre — it just
has to be a representative set of patient profiles in the same
covariate space your centres' models accept.

### Step 2: turn each centre's model into a row of `F_hat`

`f_hat_from_models()` calls `predict()` on each model at the shared grid
and stacks the rows. It dispatches on common model classes:

- **`ranger`**: extracts `predict(model, data = grid)$predictions`
- **`grf::causal_forest`** / **`regression_forest`**: extracts
  `predict(model, newdata = grid)$predictions`
- Anything else: falls through to `predict(model, newdata = grid)`

For example, with `ranger` (not run here to avoid the dependency):

```r
library(ranger)
centre_models <- lapply(centre_data_list,
                        function(d) ranger(y ~ ., data = d))
F_hat <- f_hat_from_models(centre_models, grid)
```

For `grf::causal_forest`:

```r
library(grf)
centre_models <- lapply(centre_data_list,
                        function(d) causal_forest(d$X, d$Y, d$W))
F_hat <- f_hat_from_models(centre_models, grid)
```

If your models need bespoke prediction (custom S4 classes, ensembles),
pass `predict_fn = function(model, grid) ...` returning a
length-`nrow(grid)` numeric vector.

### A worked example with `lm`

To keep this vignette dependency-free, here is the full onramp using
`lm` as the per-centre model:

```{r lm-onramp}
m <- 8
centre_meta <- data.frame(
  region    = factor(sample(c("N", "S", "E", "W"), m, replace = TRUE)),
  mean_age  = round(runif(m, 50, 70)),
  pct_female = round(runif(m, 0.4, 0.6), 2)
)

# Each centre fits a quadratic on a single covariate `x`.
make_centre_data <- function(i) {
  x <- runif(80, -1, 1)
  beta <- centre_meta$mean_age[i] / 60        # toy effect of metadata
  data.frame(x = x, y = beta * x + 0.3 * x^2 + rnorm(80, sd = 0.2))
}
centre_models <- lapply(seq_len(m), function(i)
  stats::lm(y ~ poly(x, 2), data = make_centre_data(i)))

# Build a shared grid in the centres' covariate space and stack predictions.
grid_centres <- data.frame(x = seq(-1, 1, length.out = 30))
F_hat_centres <- f_hat_from_models(centre_models, grid_centres)
dim(F_hat_centres)   # 8 × 30
```

You now have `F_hat_centres` (`m × G`) and `centre_meta` (`m × p`),
which are everything the package needs.

## A note on `grid_weights` and the L²(μ) norm

MetaHunt computes distances between functions using a
**weighted average over the grid**:

  `||f - h||² = Σ_j  weight[j] × (f(x_j) - h(x_j))²`

where `x_j` is the `j`-th grid point. The argument `grid_weights` (in
many functions) is that vector of weights — one number per grid
point. Plain English:

- **Default** (`grid_weights = NULL`): every grid point counts equally
  (`weight = 1/G`). This is fine when your grid is itself a
  representative sample of patients.
- **Non-uniform**: use this if your grid was sampled from one
  population (e.g. the source pool) but you want distances to reflect
  a different population (e.g. the target). Pass weights proportional
  to the target's density at each grid point `x_j`.

If you're not sure, leave it at the default and make sure the grid is
a representative sample of the population you care about.

## A simulated example for the rest of the vignette

For the remaining sections we'll use a *simulated* `(F_hat, W)` so that
we know the truth. **In your own data, replace this simulation with
the onramp above.**

```{r simulate}
G <- 40; m <- 120; K_true <- 3
x <- seq(0, 1, length.out = G)
basis <- rbind(sin(pi * x), cos(pi * x), x)      # 3 true bases on the grid

W <- data.frame(w1 = rnorm(m), w2 = rnorm(m))    # study-level covariates
beta <- cbind(c(1, -0.8), c(-0.5, 1.2), c(0, 0))
pi_true <- exp(as.matrix(W) %*% beta)
pi_true <- pi_true / rowSums(pi_true)

F_hat <- pi_true %*% basis + matrix(rnorm(m * G, sd = 0.05), m, G)
dim(F_hat)
```

In real data, `F_hat[i, ]` would be `predict(model_i, newdata = grid)`
and `W[i, ]` would be that centre's metadata.

## Choosing `K`

The two diagnostics in the package are complementary:

The **elbow** plot (unsupervised) shows the average projection
residual against `K`:

```{r elbow}
elbow <- reconstruction_error_curve(F_hat, K_range = 2:6,
                                    dfspa_args = list(denoise = FALSE))
plot(elbow$K, elbow$error, type = "b",
     xlab = "K", ylab = "reconstruction error")
```

Note: we set `denoise = FALSE` in this simulated example so that no
studies are pruned; in real data with noisy estimates, the default
`denoise = TRUE` is recommended.

The **CV prediction-error** plot (supervised) uses study-level
covariates:

```{r cv}
cv <- cv_error_curve(F_hat, W, K_range = 2:6, n_folds = 5,
                     dfspa_args = list(denoise = FALSE), seed = 1)
plot(cv$K, cv$cv_error, type = "b",
     xlab = "K", ylab = "CV prediction error")
```

Both dip sharply at `K = 3` — the true rank.

> **Small-`m` note.** With `m = 8` centres, `n_folds = 5` leaves only
> ~6 training studies per fold; the CV curve will be very noisy and
> the elbow may be ambiguous. Prefer `n_folds = 2` or `3`, and lean on
> domain knowledge to pick `K`.

## Tuning the denoising parameters

`dfspa()` has two denoising knobs `(N, Delta)` controlling the
neighborhood-averaging step:

```{r dfspa-knobs}
# bypass denoising entirely
fit_no <- metahunt(F_hat, W, K = 3, dfspa_args = list(denoise = FALSE))

# pass custom (N, Delta)
fit_manual <- metahunt(F_hat, W, K = 3,
                       dfspa_args = list(N = 0.5 * log(nrow(F_hat)),
                                         Delta = 0.4))
```

Or pick `(N, Delta)` by cross-validation at fixed `K`:

```{r select-denoising}
tune <- select_denoising_params(F_hat, W, K = 3, n_folds = 4, seed = 1)
tune$best
```

Defaults search `N ∈ {0.2, 0.5, 1.0} × log(m)` and
`Delta ∈ {0.05, 0.10, 0.20, 0.30} × max_pairwise_dist`. Override with
`N_grid` / `Delta_grid`.

## Fit the full pipeline

`metahunt()` chains basis hunting, constrained projection, and the
Dirichlet weight model. We plug the tuned `(N, Delta)` from the
previous section into `dfspa_args`:

```{r refit-tuned}
fit_tuned <- metahunt(F_hat, W, K = 3,
                      dfspa_args = list(N = tune$best$N,
                                        Delta = tune$best$Delta))
fit_tuned
```

We use `fit_tuned` for downstream prediction and conformal inference.

Plot the recovered bases — these are the "fundamental shapes" your
study-level functions are built from:

```{r plot-bases, fig.width = 6, fig.height = 4}
plot(fit_tuned, x_axis = x)
```

## Predict for new study-level covariates

```{r predict}
W_new <- data.frame(w1 = c(0, 1, -1), w2 = c(0, -0.5, 1))
f_pred <- predict(fit_tuned, newdata = W_new)
dim(f_pred)    # 3 x G

matplot(x, t(f_pred), type = "l", lty = 1,
        xlab = "x", ylab = expression(tilde(f)(x)),
        main = "Predicted target functions")
legend("topright", legend = paste("target", 1:3),
       col = 1:3, lty = 1, bty = "n")
```

Pass a `wrapper` for a scalar summary per target (e.g. an ATE):

```{r wrapped}
predict(fit_tuned, newdata = W_new, wrapper = mean)
```

## Conformal prediction intervals

MetaHunt offers three flavours, all returning an object you can plot
directly with `plot()`.

### Pointwise intervals (default)

With no `wrapper`, `split_conformal()` returns a separate
$(1-\alpha)$ interval **at every grid point**:

```{r split-pointwise, fig.width = 6, fig.height = 4}
res_pw <- split_conformal(F_hat, W, W_new, K = 3, alpha = 0.05, seed = 1)
plot(res_pw, target_idx = 1, x_axis = x)
```

> **Coverage caveat — pointwise vs joint.** A pointwise band gives a
> 95% interval at *each individual grid point*, not jointly across all
> grid points. The probability that the true function lies inside the
> band at *every* grid point simultaneously is smaller than 95%.
> If you need a joint guarantee, treat the band as a visualization aid
> and use a scalar wrapper for inferential claims.

### Scalar intervals via a wrapper

```{r split-scalar}
res_scalar <- split_conformal(F_hat, W, W_new, K = 3,
                              wrapper = mean, alpha = 0.05, seed = 1)
data.frame(prediction = res_scalar$prediction,
           lower      = res_scalar$lower,
           upper      = res_scalar$upper)
plot(res_scalar)
```

### Choosing a conformal variant

| Function | When to use |
|---|---|
| `split_conformal()` | Default. One train/calibration split. Fastest; some variance from the random split. |
| `cross_conformal()` | Many studies + reduce split-induced variance. Refits the pipeline `n_folds + 1` times. |
| `conformal_from_fit()` | You've already fit a pipeline (e.g. after tuning `K`) and want intervals without refitting. |

> **Small-`m` warning on `cal_frac`.** With few studies, `cal_frac = 0.3`
> can leave too few calibration sites for the chosen `alpha`. The
> finite-sample formula needs `n_cal >= ceiling((1 - alpha)(n_cal + 1))`
> calibration studies; otherwise the conformal quantile is `Inf` and
> intervals are unbounded. The package will warn when this happens.
> Either raise `cal_frac`, raise `alpha`, or use `cross_conformal()`.

### Cross-conformal example

```{r cross}
res_cross <- cross_conformal(F_hat, W, W_new, K = 3, n_folds = 5,
                             wrapper = mean, seed = 1)
res_cross
```

### Using a pre-fit pipeline

If you've already fit a `metahunt()` (e.g. after tuning `K` or
denoising parameters), use `conformal_from_fit()` with a calibration
set that was *not* used in fitting:

```{r from-fit}
tr  <- 1:80
cal <- 81:100
new <- 101:120

fit_tr <- metahunt(F_hat[tr, ], W[tr, , drop = FALSE], K = 3)

res_prefit <- conformal_from_fit(
  fit_tr$dfspa_fit, fit_tr$weight_model,
  F_cal = F_hat[cal, , drop = FALSE],
  W_cal = W[cal, , drop = FALSE],
  W_new = W[new, , drop = FALSE],
  wrapper = mean, alpha = 0.05
)

# empirical coverage on the held-out set
obs <- apply_wrapper(F_hat[new, , drop = FALSE], wrapper = mean)
mean(obs >= res_prefit$lower & obs <= res_prefit$upper)
```

## When to use `metahunt()` vs `minmax_regret()`

The package exports a second top-level method, `minmax_regret()`
(Zhang, Huang, & Imai 2024), which does **not** use study-level
covariates. It treats each source-site function as a candidate target
and computes the convex combination that minimises worst-case regret.

| Use `metahunt()` | Use `minmax_regret()` |
|---|---|
| You have many studies (`m` ≳ 20) | Few studies and/or unreliable metadata |
| Study-level covariates `W` are informative | No `W`, or `W` doesn't predict heterogeneity |
| You want predictions for **specific** new populations characterized by `W_0` | You want a single **worst-case** aggregator across the source pool |
| Random effects driven by latent structure | Adversarial / robustness framing |

```{r minmax}
mm <- minmax_regret(F_hat)
mm
```

```{r minmax-vs-metahunt, fig.width = 6, fig.height = 4}
plot(x, mm$prediction, type = "l", col = "tomato", lwd = 2,
     ylim = range(c(mm$prediction, f_pred[1, ])),
     xlab = "x", ylab = "",
     main = "minimax-regret vs. MetaHunt (target 1)")
lines(x, f_pred[1, ], col = "steelblue", lwd = 2)
legend("topright", legend = c("minmax_regret", "metahunt (W = (0,0))"),
       col = c("tomato", "steelblue"), lty = 1, lwd = 2, bty = "n")
```

## Where next

See `?metahunt`, `?split_conformal`, `?cross_conformal`,
`?conformal_from_fit`, `?select_denoising_params`, `?minmax_regret`,
`?f_hat_from_models`, `?build_grid` for full argument documentation.
The individual pipeline steps — `dfspa()`, `project_to_simplex()`,
`fit_weight_model()`, `predict_target()`,
`reconstruction_error_curve()`, `cv_error_curve()` — are all exported
and can be composed independently.

## References

Shi, W., Imai, K., and Zhang, Y. (2024). *Privacy-preserving
meta-analysis through low-rank basis hunting.*

Zhang, Y., Huang, M., and Imai, K. (2024). *Minimax regret estimation
for generalizing heterogeneous treatment effects with multisite data.*
arXiv:2412.11136.
